Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-10 paper.tex
<Enter short summary of changes>

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}
\newcommand{\sm}{\rmn{sm}}
\newcommand{\mpc}{\rmn{Mpc}}
\newcommand{\kpc}{\rmn{kpc}}
\newcommand{\mev}{\rmn{MeV}}
\newcommand{\cm}{\rmn{cm}}
%OLD
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\gev}{\rmn{GeV}}
\newcommand{\tev}{\rmn{TeV}}
\newcommand{\ev}{\rmn{eV}}
\newcommand{\dd}{\rmn{d}}
\newcommand{\mx}{\ensuremath{m_{\chi}}}
\newcommand{\ga}{\gamma}
\newcommand{\ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\ngammaj}{\ensuremath{N_{\gamma,j}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\ic}{\rmn{IC}}
\newcommand{\egamma}{\ensuremath{E_{\gamma}}}
\newcommand{\dic}{\rmn{dIC}}
\newcommand{\fsr}{\rmn{FSR}}
\newcommand{\CR}{\rmn{CR}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\rvir}{r_{200}}
\newcommand{\mvir}{M_{200}}
\newcommand{\rhoc}{\ensuremath{\rho_c}}
\newcommand{\e}{\rmn{e}}
\newcommand{\eg}{E_\gamma}
\newcommand{\ir}{\rmn{ir}}

@@

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus one the leptophilic and supersymmetric DM models
used in this work, and outline the framework for estimating the
gamma-ray emission from various radiative processes. In Section 3
we calculate the gamma-ray fluxes and spectral distributions from four
clusters and compare it to upper limits set by Fermi-LAT. For the same
clusters we derive the gamma-ray surface brightness profiles in
Section 4. To extend the analysis to more clusters and find
promising targets we estimate the fluxes from DM and CRs of all
clusters in the extended HIFLUGCS sample in Section 5. In Section
6 we conclude and discuss our results.

% --- section:  --- %
\section{Theory}
\label{sect:theory}
We start the section by discussing two different, but well motivated
DM models; leptophilic and supersymmetric DM. We then present the
framework that is used to calculate the gamma-ray emission for these
DM models using an Einasto DM density profile and the expected
enhancement from DM substructures. This is followed by an outline of
the framework of IC emission where we take into account photon fields
of the cosmic microwave background (CMB), dust, and starlight. We end
by summarizing our formalism of calculating the CR-induced pion decay
gamma-ray emission which is thought to dominate the astrophysical
gamma-ray signal.

@@

As leptophilic models, apart from being slightly contrived from the
particle physics point of view, are also rather limited by several
sets of astrophysical data, we take advantage of the recent reanalysis
\cite{Finkbeiner:2010sm} to define a benchmark leptophilic model that
is still viable.  It is found that the bounds from WMAP5 approximately
imply a Sommerfeld boost factor $S$ which has to satisfy $S(v\to
0)\lesssim 200/f\ (m_Y/1$ GeV$)$, where $f$ is the fraction of energy
from DM annihilation that ionizes the intergalactic medium,
$f\sim 0.7$ for annihilation to electrons and a factor of a few smaller
for all other standard model final states.

Taking into account the whole cosmic history of leptophilic models,
also utilizing limit on spectral and polarization distortions of the
CMB, the maximal value for the boost factor for a 1 to 2 TeV particle
and sub-GeV force carriers is found to be \cite{Finkbeiner:2010sm}
between 400 and 800. Although lower than the first estimates (e.g.,
\cite{Bergstrom:2009fa,Meade:2009iu}) this is still enough to explain
the PAMELA and Fermi excess, given other astrophysical uncertainties.

To connect easily to our previous work \cite{2009PhRvL.103r1302P}, we
choose as our benchmark leptophilic model one with mass $1.6$~TeV and
Sommerfeld boost $\B_\sfe(\sigv=200\,km/s)\approx 300$ for a Milky Way
halo and scale this appropriately down for cluster halos, using
\begin{equation}
\B_\sfe(\sigv) = 0.2\, \frac{c}{\sigv} = 
49.4 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
where the velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. Clusters have a larger velocity dispersion
and an accordingly smaller Sommerfeld boost factor. Note that this
$1/v$-scaling is valid only for $(v/c) \gtrsim (m_\phi/m_\chi)$; at
smaller velocities and outside resonances, the $1/v$-enhancement
saturates at $m_\phi/m_\chi$ \cite{2008PhRvL.101z1301K}.

@@

\subsubsection{Final state radiation}
There are two types of radiative processes which are important for our
benchmark models, first the usual QED final state radiation (FSR) for
both the leptophilic model and for supersymmetric models with charged
final states. In addition, there may be direct emission from a
virtual, charged, exchanged particle (such as a spin-0 slepton or
squark), internal bremsstrahlung, IB. The latter is essentially only
important when there is a helicity suppression for the lowest order
annihilation process, which is the case for neutralinos, whose
Majorana nature make the annihilation rate in the $s$ wave
proportional to $m_f^2$, for final state fermion $f$. Interestingly,
both the gamma-ray and fermion energy spectra are peaked at the
highest energy for this process, which can in some models cause a
rather spectacular bump in the spectra for these models.

@@

The differential energy spectrum for the IB process is more
complicated (see \cite{1989PhLB..225..372B,2008JHEP...01..049B}), but
can be computed using {\sc DarkSUSY}. The differential photon energy
spectra for all our benchmark models are shown in Fig.~\ref{fig:q_DM}.
The peaking at high photon energy and the resulting bump caused by
inner bremsstrahlung is clearly seen for benchmark models $I´$, $J´$,
and $J^*$. Despite the different particle masses and cross sections,
the spectrum of the continuums emission shows a remarkably similar
shape which suggests that a simple rescaling of particle masses
(responsible for the exponential cutoff in the spectrum) and cross
sections (spectral amplitude) should yield a roughly scale-invariant
continuum emission spectrum except for the presence of the final state
emission feature which depends on the details of the specific decay
channels. The right panel of Fig.~\ref{fig:q_DM} shows the cumulative
number of leptons (electrons and positrons) above a given energy for
our different DM models which is of interested for the high-energy IC
emission of those leptons.  Only leptophilic (LP) models have
energetic enough electrons such that the IC emission is powerful
enough to either be constrained or detected at GeV energies and
higher.

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/fit.ds.flux.eps}
 \includegraphics[width=0.49\columnwidth]{figures/fit.epflux.int.eps}
 \caption{\it Source functions for different dark matter (DM)
   models. Crosses show the simulated data from DarkSUSY and the solid
   lines show the fit to the data. LEFT PANEL: normalized differential
   continuum spectra for four different DM benchmark (BM) models;
   $\Ip$ model (red), $\Jp$ (orange), $\Kp$ (green), and $\Js$
   (blue). We use Eq.~(\ref{eq:bm_cont}) to fit the continuum
   spectra. RIGHT PANEL: number of electron and positron per DM
   annihilation above the electron energy $\ee$ for different DM
   models; $\Ip$ BM model (red), $\Jp$ BM model (orange), $\Kp$ BM
   model (green), and $\Js$ BM model (blue), leptophilic (LP) DM
   annihilating into electrons and positrons (purple), and LP DM
   annihilating into muons (black) without considering Sommerfeld
   boosts. We use Eq.~(\ref{eq:bm_elec}) to fit the spectra of
   electrons and positrons.}
 \label{fig:q_DM}
\end{minipage}
\end{figure*}

@@

In Fig.~\ref{fig:SD_spectra} we characterize the spectral shape
through a fit to the galactic spectra presented in
\cite{2006ApJ...648L..29P}.  The figure is showing the spectra at
$r=0.03\rvir$, which is the radius there the SD energy density of a
galaxy cluster equals the energy density of the CMB black-body
distribution. Inside this central radius the SD component is
dominating, which is shown in Fig.~\ref{fig:SD_Edens} where we compare
the energy densities from different radiation fields in a galaxy
cluster with the mass $\mvir=10^{15}\msun$. For this figure we use two
different profiles for the SD energy density, where the total profile
includes the contribution from the ICL, the BCG and all the galaxies,
while the galaxies are excluded in the smooth profile. To compute the
IC emission from SD, we require a non-negligible overlap of the
relativistic lepton distribution resulting form DM annihilation and
SD.  In fact, one can show with a simple order of magnitude
calculation that the overlap, $f_\rmn{IC-ol}$, of the photon field of individual
galaxies (starlight and dust emission) and the smooth DM density is
very small so that we can neglect the starlight contribution from
galaxies to the IC emission for the reminder of this work,
\begin{eqnarray}
f_\rmn{IC-ol} = \frac{N_\rmn{gal} V_\rmn{gal} f_\rmn{light}}{V_\rmn{clu}}
\lesssim  \frac{N_\rmn{gal} M_\rmn{gal}}{M_\rmn{clu}\,c_\rmn{gal}^{3}}=10^{-4}.
\end{eqnarray}
Here we assume that the exponential scale height of the stellar light
is less than the scale radius of the galaxy halo, $f_\rmn{light}
\lesssim (r_\rmn{s} / r_{200,~\rmn{gal}})^3 \sim c_\rmn{gal}^{-3} \sim
10^{-3}$ and $N_\rmn{gal} M_\rmn{gal} \sim 0.1 M_\rmn{clu}$.

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/fit.porter.v2.eps}
 \caption{\it Spectral dependence of radiation fields in a cluster of
   galaxies. The black dotted line in the left peak show the spectrum
   of CMB photons using a black body with a temperature of
   $2.73$~K. The crosses in the middle and right peaks represent the
   measured spectra from stars extending from the near IR to UV and
   dust at far IR wavelengths (SD), respectively, and are derived in
   \cite{2006ApJ...648L..29P} for a galaxy. We normalize the
   individual SD spectrum separately using the observed luminosity
   from SD in clusters. The SD luminosity is related to the cluster
   mass through Eqs.~(\ref{eq:u_SD}-\ref{eq:N_stars}), where we use
   have used the mass $\mvir=10^{15}\msun$ in this figure. We
   renormalize the SD spectra to the radius $r=0.03\rvir$, where the
   smooth energy density of the SD light (see Fig.~\ref{fig:SD_Edens})
   equals the energy density of the CMB. The red dashed lines show the
   fitted SD spectral model.}
 \label{fig:SD_spectra}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/ucool.eps}
 \caption{\it Spatial dependence of the energy density of radiation
   fields in a cluster of galaxies. The energy density of the CMB
   (blue dotted line) is isotropic with
   $u_\rmn{cmb}=0.26\,\rmn{eV}\,\rmn{cm}^{-3}$. The energy density of
   the light from stars and dust (SD) is denoted by the red dashed
   line and the solid purple line for the total SD light and the
   smooth SD light, respectively. For comparison we show the energy
   density of the stars and a low dust model (SLD) with the light blue
   line. The SD light has been renormalized to a cluster with the mass
   $\mvir=10^{15}\msun$. Finally we show the energy density two
   magnetic field models with a central magnetic field of 3~$\mu
   G$. The magnetic field scales with the gas density to the power
   $\alpha_\rmn{B}$; green dash-dotted line ($\alpha_\rmn{B}=0.5$) and
   orange dash-dotted line ($\alpha_\rmn{B}=0.7$). Note that the SD
   radiation is dominating the energy density inside
   $\sim0.1\,\rvir$.}
 \label{fig:SD_Edens}
\end{figure}

@@

First we focus on gamma-ray flux-cluster mass scaling relations for DM
annihilation and CR induced emission in Fig.~\ref{fig:lum_mass_scaling}.  The
mass scaling of the substructure boost steepens the intrinsically shallower DM
annihilation relation of benchmark models (Eq.~\ref{eq:BM_scaling}) while
this steepening is in fact over-compensated by the inverse mass scaling of the
Sommerfeld boost in leptophilic models that prefers smaller halos with a larger
ratio of $c/v_\mathrm{circ}$ (Eq.~\ref{eq:LP_scaling}). In contrast, the
CR scaling relation is considerably steeper as shown by reference
\cite{2010MNRAS.409..449P}: 
\begin{eqnarray}
L_{\rmn{CR}-\ensuremath{\pi^0}}(>100\,\mev) = 3.1\times10^{45}
\left(\frac{\mvir}{10^{15}\,\msun}\right)^{1.46}\mbox{erg~s}^{-1}\nonumber\\
L_{\rmn{CR}-\ensuremath{\pi^0}}(>10\,\gev)  = 2.2\times10^{43}
\left(\frac{\mvir}{10^{15}\,\msun}\right)^{1.34}\mbox{erg~s}^{-1}\nonumber\\
\end{eqnarray}
This implies already a general strategy to minimize the CR-induced
foreground for DM annihilation and argues for very nearby groups. We
note that these simulations do not include AGN feedback that is
thought to furthermore reduce the baryon fraction in groups relative
to that in clusters \cite{2008ApJ...687L..53P}. The resulting smaller
target density for hadronic CR interactions will furthermore steepen
the $L_\gamma-M$ relation of CR induced emission, making the case for
groups even stronger.

@@

\section{Detailed modelling of  starlight and dust in clusters}

\subsection{Energy spectrum}

Our goal in this section is to derive a simple spectral model for SD
in galaxy clusters. The SD light from clusters is dominated by the
emission from galaxies. This radiation is expected to leak out into
the much larger volume of the ICM, hence we approximate the spectral
distribution of the ICL with the light from galaxies. We use the far
infrared to ultraviolet spectrum derived in \cite{2006ApJ...648L..29P}
for a MW type galaxy to characterize the SD spectral distribution of a
cluster. Note that we only keep the spectral shape, and renormalize
the spectral distribution using the luminosity from SD.

In Fig.~\ref{fig:SD_spectra} we show the spectral distribution of the
light from SD together with the CMB black-body distribution in a
galaxy cluster, both at the radius $r=0.02\rvir$ where the energy
density of the smooth light from SD (see Fig.~\ref{fig:SD_Edens})
equals the energy density of the CMB. We have renormalized both the
spectral data and the fitted spectra using the luminosity from stars
and dust individually.

We choose to fit the spectrum of the dust that peaks at about
$10^{-2}\,\ev$ with a double power-law, and the slightly wider
spectral distribution of the stars that peaks at about $1\,\ev$ with a
tipple power-law. We find the best fit spectra for the SD in a
cluster through:
\begin{eqnarray}
  \eph^2\frac{\dd^2 N_\ph}{\dd \eph \dd V}
  &\equiv& u_\sd(\eph) =  \sum_i N_i(\mvir) u_i^\gal(\eph)\,,\nonumber \\ 
\rmn{where}\quad i&=&\{\rmn{stars,dust}\} \qquad \rmn{and}\\ 
% &=&  N_\dust(\lx) u_\dust^\gal(\eph) + N_\stars(\lx) u_\stars^\gal(\eph)  \nonumber \\ 
%  \\
  u_\stars^\gal(\eph) &=& \frac{23\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{1.23\,\rmn{eV}}{\eph}\right)^{1.9} \nonumber \\
  &\times&\left[1+\left(\frac{2.04\,\rmn{eV}}{\eph}\right)^{20}\right]
  ^{-\frac{1.9}{20}}\nonumber \\
  &\times& \left[1+\left(\frac{0.78\,\rmn{eV}}{\eph}\right)^{20}\right]^{-\frac{2.6}{20}}\,, \\
  u_\dust^\gal(\eph) &=& 
  \frac{40\,\rmn{eV}}{\rmn{cm}^3} 
  \left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{4.9}\nonumber \\
  &\times& \left[1+\left(\frac{0.0144\,\rmn{eV}}{\eph}\right)^{1.9}\right]^{-4.9}\,.
\end{eqnarray}

>>>>

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}
\newcommand{\sm}{\rmn{sm}}
\newcommand{\mpc}{\rmn{Mpc}}
\newcommand{\kpc}{\rmn{kpc}}
\newcommand{\mev}{\rmn{MeV}}
\newcommand{\cm}{\rmn{cm}}
\newcommand{\egt}{\tilde{E}_\gamma}
\newcommand{\eet}{\tilde{E}_\rmn{e}}
%OLD
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\gev}{\rmn{GeV}}
\newcommand{\tev}{\rmn{TeV}}
\newcommand{\ev}{\rmn{eV}}
\newcommand{\dd}{\rmn{d}}
\newcommand{\mx}{\ensuremath{m_{\chi}}}
\newcommand{\ga}{\gamma}
\newcommand{\ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\ngammaj}{\ensuremath{N_{\gamma,j}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\ic}{\rmn{IC}}
\newcommand{\egamma}{\ensuremath{E_{\gamma}}}
\newcommand{\dic}{\rmn{dIC}}
\newcommand{\fsr}{\rmn{FSR}}
\newcommand{\CR}{\rmn{CR}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\rvir}{r_{200}}
\newcommand{\mvir}{M_{200}}
\newcommand{\rhoc}{\ensuremath{\rho_c}}
\newcommand{\e}{\rmn{e}}
\newcommand{\eg}{E_\gamma}
\newcommand{\ir}{\rmn{ir}}

@@

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus one the LP and supersymmetric DM models
used in this work, and outline the framework for estimating the
gamma-ray emission from various radiative processes. In Section 3
we calculate the gamma-ray fluxes and spectral distributions from four
clusters and compare it to upper limits set by Fermi-LAT. For the same
clusters we derive the gamma-ray surface brightness profiles in
Section 4. To extend the analysis to more clusters and find
promising targets we estimate the fluxes from DM and CRs of all
clusters in the extended HIFLUGCS sample in Section 5. In Section
6 we conclude and discuss our results.

% --- section:  --- %
\section{Theory}
\label{sect:theory}
We start the section by discussing two different, but well motivated
DM models; LP and supersymmetric DM. We then present the framework
that is used to calculate the gamma-ray emission for these DM models
using an Einasto DM density profile and the expected enhancement from
DM substructures. This is followed by an outline of the framework of
IC emission where we take into account photon fields of the cosmic
microwave background (CMB), dust, and starlight. We end by summarizing
our formalism of calculating the CR-induced pion decay gamma-ray
emission which is thought to dominate the astrophysical gamma-ray
signal.

@@

As leptophilic models, apart from being slightly contrived from the
particle physics point of view, are also rather limited by several
sets of astrophysical data, we take advantage of the recent reanalysis
\cite{Finkbeiner:2010sm} to define a benchmark LP model that
is still viable.  It is found that the bounds from WMAP5 approximately
imply a Sommerfeld boost factor $S$ which has to satisfy $S(v\to
0)\lesssim 200/f\ (m_Y/1$ GeV$)$, where $f$ is the fraction of energy
from DM annihilation that ionizes the intergalactic medium,
$f\sim 0.7$ for annihilation to electrons and a factor of a few smaller
for all other standard model final states.

Taking into account the whole cosmic history of LP models, also
utilizing limit on spectral and polarization distortions of the CMB,
the maximal value for the boost factor for a 1 to 2~TeV particle and
sub-GeV force carriers is found to be \cite{Finkbeiner:2010sm} between
400 and 800. Although lower than the first estimates (e.g.,
\cite{Bergstrom:2009fa,Meade:2009iu}) this is still enough to explain
the PAMELA and Fermi excess, given other astrophysical uncertainties.

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of ($1/2,\,1/2$)
into ($\mu^+/\mu^-,\,\e^+/\e^-$) \footnote{This is an approximation to
  the model found in \cite{Finkbeiner:2010sm} that has a branching
  ratio of ($1/3,\,1/3,\,1/3$) into
  ($\mu^+/\mu^-,\,\e^+/\e^-,\,\pi^+/\pi^-$).}. The Sommerfeld boost
$S=\B_\sfe(\sigv=200\,km/s)\approx 300$ for a Milky Way halo in our LP
model, which we appropriately scale down for cluster halos, using
\del{To connect easily to our previous work
  \cite{2009PhRvL.103r1302P}, we choose as our benchmark LP model one
  with mass $1.6$~TeV and Sommerfeld boost
  $S=\B_\sfe(\sigv=200\,km/s)\approx 300$ for a Milky Way halo and
  scale this appropriately down for cluster halos, using} {\bf Anders:
  insert Sommerfeld boost formula as a function of halo mass, done}
\begin{equation}
\B_\sfe(\sigv) = 0.2\, \frac{c}{\sigv} = 
49.4 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
where the velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. Clusters have a larger velocity dispersion
and an accordingly smaller Sommerfeld boost factor. Note that this
$1/v$-scaling is valid only for $(v/c) \gtrsim (m_\phi/m_\chi)$; at
smaller velocities and outside resonances, the $1/v$-enhancement
saturates at $m_\phi/m_\chi$ \cite{2008PhRvL.101z1301K}.

@@

\subsubsection{Final state radiation}
There are two types of radiative processes which are important for our
benchmark models, first the usual QED final state radiation (FSR) for
both the LP model and for supersymmetric models with charged final
states. In addition, there may be direct emission from a virtual,
charged, exchanged particle (such as a spin-0 slepton or squark),
internal bremsstrahlung, IB. The latter is essentially only important
when there is a helicity suppression for the lowest order annihilation
process, which is the case for neutralinos, whose Majorana nature make
the annihilation rate in the $s$ wave proportional to $m_f^2$, for
final state fermion $f$. Interestingly, both the gamma-ray and fermion
energy spectra are peaked at the highest energy for this process,
which can in some models cause a rather spectacular bump in the
spectra for these models.

@@

The differential energy spectrum for the IB process is more
complicated (see \cite{1989PhLB..225..372B,2008JHEP...01..049B}), but
can be computed using {\sc DarkSUSY}. The differential photon energy
spectra for all our benchmark models are shown in Fig.~\ref{fig:q_DM}.
The peaking at high photon energy and the resulting bump caused by
inner bremsstrahlung is clearly seen for benchmark models $I´$, $J´$,
and $J^*$. Despite the different particle masses and cross sections,
the spectrum of the continuums emission shows a remarkably similar
shape which suggests that a simple rescaling of particle masses
(responsible for the exponential cutoff in the spectrum) and cross
sections (spectral amplitude) should yield a roughly scale-invariant
continuum emission spectrum except for the presence of the final state
emission feature which depends on the details of the specific decay
channels. The right panel of Fig.~\ref{fig:q_DM} shows the cumulative
number of leptons (electrons and positrons) above a given energy for
our different DM models which is of interested for the high-energy IC
emission of those leptons. Only LP models have energetic enough
electrons such that the IC emission is powerful enough to either be
constrained or detected at GeV energies and higher.

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/fit.ds.flux.eps}
 \includegraphics[width=0.49\columnwidth]{figures/fit.epflux.int.eps}
 \caption{\it Source functions for different dark matter (DM)
   models. Crosses show the simulated data from {\sc DarkSUSY} and the
   solid lines show the fit to the data. LEFT PANEL: normalized
   differential continuum spectra for four different DM benchmark (BM)
   models; $\Ip$ model (red), $\Jp$ (orange), $\Kp$ (green), and $\Js$
   (blue). We use Eq.~(\ref{eq:bm_cont}) to fit the continuum
   spectra. RIGHT PANEL: number of electron and positron per DM
   annihilation above the electron energy $\ee$ for different DM
   models; $\Ip$ BM model (red), $\Jp$ BM model (orange), $\Kp$ BM
   model (green), and $\Js$ BM model (blue), leptophilic (LP) DM
   annihilating into electrons and positrons (purple), and LP DM
   annihilating into muons (black) without considering Sommerfeld
   boosts. We use Eq.~(\ref{eq:bm_elec}) to fit the spectra of
   electrons and positrons.}
 \label{fig:q_DM}
\end{minipage}
\end{figure*}

@@

In Fig.~\ref{fig:SD_spectra} we characterize the spectral shape
through a fit to the galactic spectra presented in
\cite{2006ApJ...648L..29P}.  The figure is showing the spectra at
$r=0.03\rvir$, which is the radius there the SD energy density of a
galaxy cluster equals the energy density of the CMB black-body
distribution. Inside this central radius the SD component is
dominating, which is shown in Fig.~\ref{fig:SD_Edens} where we compare
the energy densities from different radiation fields in a galaxy
cluster with the mass $\mvir=6\times10^{14}\msun$. For this figure we
use two different profiles for the SD energy density, where the total
profile includes the contribution from the ICL, the BCG and all the
galaxies, while the galaxies are excluded in the smooth profile. To
compute the IC emission from SD, we require a non-negligible overlap
of the relativistic lepton distribution resulting form DM annihilation
and SD.  In fact, one can show with a simple order of magnitude
calculation that the overlap, $f_\rmn{IC-ol}$, of the photon field of
individual galaxies (starlight and dust emission) and the smooth DM
density is very small so that we can neglect the starlight
contribution from galaxies to the IC emission for the reminder of this
work,
\begin{eqnarray}
f_\rmn{IC-ol} = \frac{N_\rmn{gal} V_\rmn{gal} f_\rmn{light}}{V_\rmn{clu}}
\lesssim  \frac{N_\rmn{gal} M_\rmn{gal}}{M_\rmn{clu}\,c_\rmn{gal}^{3}}=10^{-4}.
\end{eqnarray}
Here we assume that the exponential scale height of the stellar light
is less than the scale radius of the galaxy halo, $f_\rmn{light}
\lesssim (r_\rmn{s} / r_{200,~\rmn{gal}})^3 \sim c_\rmn{gal}^{-3} \sim
10^{-3}$ and $N_\rmn{gal} M_\rmn{gal} \sim 0.1 M_\rmn{clu}$.

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/fit.porter.v2.eps}
 \caption{\it Spectral dependence of radiation fields in a cluster of
   galaxies. The black dotted line in the left peak show the spectrum
   of CMB photons using a black body with a temperature of
   $2.73$~K. The crosses in the middle and right peaks represent the
   measured spectra from stars extending from the near IR to UV and
   dust at far IR wavelengths (SD), respectively, and are derived in
   \cite{2006ApJ...648L..29P} for a galaxy. We normalize the
   individual SD spectrum separately using the observed luminosity
   from SD in clusters. The SD luminosity is related to the cluster
   mass through Eqs.~(\ref{eq:E_SD}-\ref{eq:N_dust}), where we use
   have used the mass $\mvir=6\times10^{14}\msun$ in this figure. We
   renormalize the SD spectra to the radius $r=0.03\rvir$, where the
   smooth energy density of the SD light (see Fig.~\ref{fig:SD_Edens})
   equals the energy density of the CMB. The red dashed lines show the
   fitted SD spectral model.}
 \label{fig:SD_spectra}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/ucool.eps}
 \caption{\it Spatial dependence of the energy density of radiation
   fields in a cluster of galaxies. The energy density of the CMB
   (blue dotted line) is isotropic with
   $u_\rmn{cmb}=0.26\,\rmn{eV}\,\rmn{cm}^{-3}$. The energy density of
   the light from stars and dust (SD) is denoted by the red dashed
   line and the solid purple line for the total SD light and the
   smooth SD light, respectively. For comparison we show the energy
   density of the stars and a low dust model (SLD) with the light blue
   line. The SD light has been renormalized to a cluster with the mass
   $\mvir=6\times10^{14}\msun$. Finally we show the energy density two
   magnetic field models with a central magnetic field of 3~$\mu
   G$. The magnetic field scales with the gas density to the power
   $\alpha_\rmn{B}$; green dash-dotted line ($\alpha_\rmn{B}=0.5$) and
   orange dash-dotted line ($\alpha_\rmn{B}=0.7$). Note that the SD
   radiation is dominating the energy density inside
   $\sim0.1\,\rvir$.}
 \label{fig:SD_Edens}
\end{figure}

@@

First we focus on gamma-ray flux-cluster mass scaling relations for DM
annihilation and CR induced emission in Fig.~\ref{fig:lum_mass_scaling}.  The
mass scaling of the substructure boost steepens the intrinsically shallower DM
annihilation relation of benchmark models (Eq.~\ref{eq:BM_scaling}) while
this steepening is in fact over-compensated by the inverse mass scaling of the
Sommerfeld boost in leptophilic models that prefers smaller halos with a larger
ratio of $c/v_\mathrm{circ}$ (Eq.~\ref{eq:LP_scaling}). In contrast, the
CR scaling relation is considerably steeper as shown by reference
\cite{2010MNRAS.409..449P}: 
\begin{eqnarray}
L_{\gamma}(>100\,\mev) = 3.1\times10^{45}
\left(\frac{\mvir}{10^{15}\,\msun}\right)^{1.46}\mbox{erg~s}^{-1}\nonumber\\
L_{\gamma}(>10\,\gev)  = 2.2\times10^{43}
\left(\frac{\mvir}{10^{15}\,\msun}\right)^{1.34}\mbox{erg~s}^{-1}\,.\nonumber\\
\end{eqnarray}
Note that the CR luminosity scaling relations include the IC gamma-ray
contribution from shock accelerated electrons as well as secondary
electrons created through CRp-proton interactions. The gamma-ray flux
from the electrons is, however, dominated by the decaying $\pi^0$ and
only show up as a small flattening of the scaling relation at energies
around $100\,\gev$. The steeper mass scaling of the CR compared to the
DM implies already a general strategy to minimize the CR-induced
foreground for DM annihilation and argues for very nearby groups. We
note that these simulations do not include AGN feedback that is
thought to furthermore reduce the baryon fraction in groups relative
to that in clusters \cite{2008ApJ...687L..53P}. The resulting smaller
target density for hadronic CR interactions will furthermore steepen
the $L_\gamma-M$ relation of CR induced emission, making the case for
groups even stronger.

@@

\section{Detailed modelling of starlight and dust in clusters}
\label{sect:SD}
In this section we show the details of our model of stars and dust
(SD) in clusters of galaxies. In the first part we use the spectral
distribution of the SD light in galaxies to derive a simple spectral
model for SD in galaxy clusters. This is done by renormalizing the
galactic SD light to match the conditions for a cluster. In the second
part we derive the spatial distirbution of the SD light.

\subsection{Energy spectrum}
The SD light from clusters is dominated by the emission from
galaxies. This radiation is expected to leak out into the much larger
volume of the ICM, hence we approximate the spectral distribution of
the ICL with the light from galaxies. We use the far infrared to
ultraviolet spectrum derived in \cite{2006ApJ...648L..29P} for a MW
type galaxy to characterize the SD spectral distribution of a
cluster. Note that we only keep the spectral shape, and renormalize
the spectral distribution using the luminosity from SD.
